Dependent Object Types

### 2012-08-30 doc/candidacy/report.tex
path-dep types in tribe
rev

Scala's abstract types have close resemblances to abstract types of
signatures in these ML-like module systems. In particular, in their
calculus, there's a form of dependent type, since translucent sums are
expressions that may contain types. Compared to path-dependent types
as found in Scala, these are more limited because recursive references
are not allowed.

@@

In~\cite{tribe}, Clarke et al. present a calculus for virtual
classes. Virtual classes allows {\it families} of classes to be
inherited, rather than just single classes. Classes are lexically
nested inside other classes. When a class is inherited, its nested
inner classes are inherited along with its methods and fields. {\it
  Further binding} enables the nested classes to be further extended.

Though Scala doesn't have virtual classes, its abstract types enables
it to emulate some (but not all) of the benefits of virtual classes as
presented in the Tribe calculus.

In terms of theoretical results, there's a sketch of soundness in the
paper, but it's unclear whether it is correct. Decidability and
completeness of subtyping has not yet been established.

>>>>

Scala's abstract types have close resemblances to abstract types of
signatures in these ML-like module systems. In particular, in their
calculus, there's a form of dependent type, since translucent sums are
expressions that may contain types. Compared to path-dependent types
as found in Scala, these are more limited because recursive references
and bounded quantifications are not allowed.

@@

In~\cite{tribe}, Clarke et al. present a calculus for virtual
classes. Virtual classes allow {\it families} of classes to be
inherited, rather than just single classes. Classes are lexically
nested inside other classes. When a class is inherited, its nested
inner classes are inherited along with its methods and fields. {\it
  Further binding} enables the nested classes to be further
extended. Path-dependent types arise naturally to distinguish objects
coming from different families. In Tribe, paths can additionally use
an ``out'' field to move from an object to the object which surrounds
it.

%Tribe, like Scala, distinguishes object families by using
%path-dependent types. An {\it object family type} such as {\code
%  x.f.C} refers to class {\code C} which is nested within the object
%at the end of path {\code x.f}. Paths have to be invariant, and so
%consists of chains of immutable fields or variables.

Though Scala doesn't have virtual classes, its abstract types enables
it to emulate some (but not all) of the benefits of virtual classes as
presented in the Tribe calculus. On the other hand, Tribe doesn't have
virtual types. In a follow-up paper~\cite{tribalo}, an improved
variant of Tribe with simpler types, cross-family inheritance and
generics is presented and a formalisation of ownership types built
upon it.

In terms of theoretical results, there's a sketch of soundness in the
paper, but it's unclear whether it is correct. In particular, the
substitution lemma does not seem valid when null is the value to be
substituted and the expression has a type which contains the variable
to be substituted. In the follow-up paper~\cite{tribalo}, the
soundness proof sketch seems more plausible but then is tied to the
ownership system formalism. Decidability and completeness of subtyping
has not yet been established.
